\documentclass[11pt,american]{article}
\usepackage[letterpaper,hmargin=1.25in,vmargin=1.25in]{geometry}

\usepackage[utf8]{inputenc}
\usepackage[mathscr]{euscript}
%\usepackage[affil-it]{authblk}
\usepackage{authblk}
\usepackage{times}
\usepackage{babel}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{verbatim} % For comment env
\usepackage{enumerate}
\usepackage{afterpage}
\usepackage{hyperref} % For internal links
\usepackage{nag}      % Warn about old packages and commands
\usepackage{fixltx2e} % Fix misc. LaTeX stuff
\usepackage{xspace}   % Fix spacing at end of macros
\usepackage{array}
\usepackage{etoolbox}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{caption}
\usepackage{float}    % 'H' table placement option
%\usepackage{compactbib} % More compact bibliography

\usepackage{tikz}
\usetikzlibrary{calc}

\setcounter{MaxMatrixCols}{16}

\newenvironment{smallpmatrix}
  {\left(\begin{smallmatrix}}
  {\end{smallmatrix}\right)}

% For now, to help organize
\setcounter{tocdepth}{5}
\newcommand{\TODO}{\textcolor{red}{\textbf{TODO}}\xspace}

\newcommand{\Ztwo}{\ensuremath{\mathbb{Z}_2}\xspace}
\newcommand{\Zn}{\ensuremath{\mathbb{Z}_n}\xspace}
\newcommand{\gftwo}{\ensuremath{\mathrm{GF}(2)}\xspace}
\newcommand{\gfsixteen}{\ensuremath{\mathrm{GF}(2^{16})}\xspace}
\newcommand{\gfwithpoly}{\ensuremath{\text{GF}(2^{16}) / \left\langle \ourpoly \right\rangle}\xspace}

\newcommand{\pval}{\textit{P-value}\xspace}
\newcommand{\pvals}{\textit{P-values}\xspace}

\newcommand{\Keccak}{\textsc{Keccak}\xspace}
\newcommand{\SpongeWrap}{\textsc{SpongeWrap}\xspace}

\newcommand{\drawxTimes}[2]{% center: x,y
  \draw[fill=gray!15,line width=2.5pt] (#1, #2) ellipse (1em and 1em) node {$x*$};
}
\newcommand{\drawXOR}[2]{% center: x,y
  \draw[fill=gray!15,line width=2.5pt] (#1, #2) ellipse (1em and 1em);
  \draw[line width=2.5pt] (#1-.4, #2) -- (#1+.4, #2);
  \draw[line width=2.5pt] (#1, #2-.4) -- (#1, #2+.4);
}
\newcommand{\drawAdder}[2]{% top left: x,y
  \draw[fill=gray!15,line width=2.5pt] (#1, #2) rectangle (#1+1, #2-1);
  \draw[line width=2.5pt] (#1, #2-.5) -- (#1+1, #2-.5);
  \draw[line width=2.5pt] (#1+.5, #2-1) -- (#1+.5, #2);
}
\newcommand{\drawRot}[2]{% center: x,y
  \draw[fill=gray!15,line width=2.5pt] (#1, #2) ellipse (2em and 1em) node {$\mathbf{ROT}$};
}
\newcommand{\drawMixerInputs}{
  \draw[fill=gray!30] (0,16) rectangle (2,15) node[midway] {$A$};
  \draw[fill=gray!30] (4,16) rectangle (6,15) node[midway] {$B$};
}
\newcommand{\drawMixerOutputs}[1]{%y offset from top
  \draw[fill=gray!30] (0,16-#1) rectangle (2,15-#1) node[midway] {$A'$};
  \draw[fill=gray!30] (4,16-#1) rectangle (6,15-#1) node[midway] {$B'$};
}

% Aliases for readability
\let\from=\colon
\let\to=\rightarrow

% TODO better title
%\title{A Customizable Sponge-Based Authenticated Encryption Algorithm with $16$-bit S-boxes}
\title{Customizable Sponge-Based Authenticated Encryption Using $16$-bit S-boxes}
%\author{Matthew Kelly
        %\and Alan Kaminsky
        %\and Marcin {\L}ukowiak
        %\and Michael Kurdziel
        %\and Stanis{\l}aw Radziszowski}
\author[1]{Matthew Kelly}
\author[2]{Alan Kaminsky}
\author[1]{Marcin {\L}ukowiak}
\author[3]{\authorcr Michael Kurdziel}
\author[2]{Stanis{\l}aw Radziszowski}
%\affil[1]{\footnotesize Department of Computer Engineering, \authorcr Rochester Institute of Technology}
\affil[1]{\footnotesize Department of Computer Engineering, Rochester Institute of Technology}
%\affil[2]{Department of Computer Science,  \authorcr Rochester Institute of Technology}
\affil[2]{Department of Computer Science, Rochester Institute of Technology}
\affil[3]{Harris Corporation, Rochester, NY}

\date{September 2014}

\begin{document}

\maketitle
%\footnote{\TODO Harris acknowledgement?}

\begin{abstract}
Authenticated encryption (AE) is a symmetric key cryptographic scheme that aims to provide both confidentiality and data integrity. 
There are many AE algorithms in existence today.
However, they are often far from ideal in terms of efficiency and ease of use.
For this reason, there is ongoing effort to develop new AE algorithms that are secure, efficient, and easy to use.

The sponge construction is a relatively new cryptographic primitive that has gained popularity since the SHA-3 hashing competition was won by the sponge-based \Keccak algorithm.
The duplex construction, which is closely related to the sponge, provides promising potential for secure and efficient authenticated encryption.

In this paper we introduce a novel authenticated encryption algorithm based on the duplex construction that is targeted for hardware implementation.
We also provide explicit customization guidelines for users who desire unique authenticated encryption solutions within our security margin.
Futhermore, our substitution step uses $16 \times 16$ AES-like S-boxes.
We believe this to be the first algorithm in the literature to use such large bijective S-boxes.
\end{abstract}

\input{tex/IntroSection.tex}
\input{tex/SpongeAndDuplexSection.tex}
\input{tex/AlgorithmSpecSection.tex}
\input{tex/CryptanalysisSection.tex}
\input{tex/ConclusionsSection.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1}
\selectfont
\bibliographystyle{ieeetr}
\bibliography{paper}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
\newpage
\input{tex/BitwisePermutationsAppendix.tex}
\input{tex/RoundConstantsAppendix.tex}

\end{document}

